Omics-based Single Step Trait Prediction

### 2017-09-17 paper/manuscript.tex
paper, fix typos

We wish to submit the enclosed manuscript "Omics-based Single Step Prediction"
by Westhues et al.\@ for publication in TAG.

@@

In conclusion our results demonstrate that single step prediction, using
an established predictor such as pedigree information, can benefit from sampling
a highly performant predictor, such as transcriptomic data, on a subset of all
genotypes.
The framework presented herein should help promoting the use of attractive
`omics' predictors by balancing their advantages with their currently higher
sampling costs.

@@

\section{Materials and Methods}
\Genetics2level{Exp1}
To address the second objective a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four subgroups classified by
\citeN{Yang2014}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
17 traits~\cite{Guo2016} of which six are presented in this study
(Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were predicted with
greater precision by transcriptomic data and two further traits were predicted
with greater precision by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

@@

Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[_{I}]{\mathbf{W}}{} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \\
  \tensor[_I]{\mathbf{W}}{_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[_I]{\mathbf{W}}{_{2}}$
while the feature covariates for genotypes not covered by the incomplete
predictor can be estimated/imputed with the aid of the covariance matrix of the
complete predictor $\tensor[_{C}]{\mathbf{K}}{}$ and the feature matrix of the
incomplete predictor $\tensor[_{I}]{\mathbf{W}}{_{22}}$ by calculating
($\tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} = \tensor[_C]{\mathbf{K}}{_{12}} \;
\tensor[_C]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_I]{\mathbf{W}}{_{2}}$).
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \tensor[_{I}]{\mathbf{W}}{_{2}} \; 
\tensor[_{I}]{\boldsymbol{\alpha}}{}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\tensor[_{I}]{\boldsymbol{\alpha}}{}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[_{C}]{\mathbf{K}}{_{12}} \;
       	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{} +
       	(\mathbf{g_1} - \tensor[_{C}]{\mathbf{K}}{_{12}} \;
	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

\Genetics2level{Impact of coverage by the incomplete predictor}
Predictive abilities for the six traits in the the different core sets of Exp1
varied widely within most traits (Fig.~\ref{fig:core-results}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 10\% to 20\%/30\% was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were better predicted via G than via T in
the reduced set of lines, predictive abilities were higher when only a small
number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted better when using T
individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).

>>>>

We wish to submit the enclosed manuscript ``Omics-based Single Step Prediction''
by Westhues et al.\@ for publication in TAG.

@@

In conclusion our results demonstrate that single step prediction, using
an established predictor such as pedigree information, can benefit from sampling
a highly performant predictor, such as transcriptomic data, on a subset of all
genotypes.
The framework presented herein should help promote the use of attractive
`omics' predictors by balancing their advantages with their currently higher
sampling costs.

@@

\section{Materials and Methods}
\Genetics2level{Exp1}
To address the second objective a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four subgroups classified by
\citeN{Guo2016}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
17 traits of which six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that two traits were predicted with
greater precision by transcriptomic data and four further traits were predicted
with greater precision by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

@@

Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[_{I}]{\mathbf{W}}{} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \\
  \tensor[_I]{\mathbf{W}}{_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[_I]{\mathbf{W}}{_{2}}$
while the feature covariates for genotypes not covered by the incomplete
predictor can be estimated/imputed with the aid of the covariance matrix of the
complete predictor $\tensor[_{C}]{\mathbf{K}}{}$ and the feature matrix of the
incomplete predictor $\tensor[_{I}]{\mathbf{W}}{_{2}}$ by calculating
($\tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} = \tensor[_C]{\mathbf{K}}{_{12}} \;
\tensor[_C]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_I]{\mathbf{W}}{_{2}}$).
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \tensor[_{I}]{\mathbf{W}}{_{2}} \; 
\tensor[_{I}]{\boldsymbol{\alpha}}{}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\tensor[_{I}]{\boldsymbol{\alpha}}{}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[_{C}]{\mathbf{K}}{_{12}} \;
       	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{} +
       	(\mathbf{g_1} - \tensor[_{C}]{\mathbf{K}}{_{12}} \;
	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

\Genetics2level{Impact of coverage by the incomplete predictor}
Predictive abilities for the six traits in the different core sets of Exp1
varied widely within most traits (Fig.~\ref{fig:core-results}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 10\% to 20\%/30\% was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were predicted with higher precision via G
than via T in the reduced set of lines, predictive abilities were higher when
only a small number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted with greater precision
when using T individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).
